Soling at the 2000 Summer Olympics

### 2020-07-29 inst/f1points.Rmd
work

\begin{multline}
\mbox{Hamilton}\succ
\mbox{Vettel}\succ
\mbox{Bottas}\succ
\mbox{Raikkonen}\succ
\mbox{Ricciardo}\succ
\mbox{Verstappen}\succ\\
\mbox{Perez1}\succ
\mbox{Ocon}\succ
\mbox{Sainz}\succ
\mbox{Hulkenberg}\succ
\mbox{Massa}
\end{multline}

@@

\begin{multline}
\mbox{Hamilton}\succ
\mbox{Vettel}\succ
\mbox{Bottas}\succ
\mbox{Ricciardo}\succ
\mbox{Verstappen}\succ
\mbox{Raikkonen}\succ\\
\mbox{Perez}\succ
\mbox{Ocon}\succ
\mbox{Massa}\succ
\mbox{Sainz}\succ
\mbox{Hulkenberg}\succ
\end{multline}

@@

```{r,label=likelihoodrankingshower,echo=FALSE}
jj <- mH2017
jj <- sort(jj,decreasing=TRUE)
jj[] <- seq_along(jj)
```

@@

We can plot one ranking against the other\footnote{The historically
correct points awarded to the drivers differs from that calculated
here.  That is for two reasons: firstly, ``fastest lap'' points are
not included here, and also the truncation of the order table to the
first 11 drivers can increase the rank of a driver if non-first-11
drivers are placed} , shown in Figure
\ref{orderingbypointsandlikelihood}.  We thus see a comparison between
the two ordering systems.  Taking 2017 as an example, drivers Hamilton
and Vettel are are respectively first and second according to both
ranking ranking procedures; but R\"{a}ikk\"{o}nen YES I KNOW IT IS
WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
fourth and third, according to points and likelihood respectively.
The degree of agreement between the two systems might be quantified as
"2", as the positions of the top two drivers agree.  Thus the degree
of agreement between likelihood ranking and points ranking for the
years 2016-9 would be 8,2,6, and 2 respectively.

However, the points system used is essentially arbitrary.  We could
use, for example, Zipf's law to rank the drivers: award one point to
the winner, half a point to second place, one third of a point to
third, and so on; see figure \ref{orderingbypointsandlikelihood_zipf}
in which we see generally poorer agreement between points-based ranks
and likelihood-based ranks.  There are a number of plausible points
systems that might be used:

@@

```{r,label=likevspointszipf,cache=TRUE,echo=FALSE,fig.cap="As for figure \ref{likevspoints} \\label{orderingbypointsandlikelihood_zipf} but points ranking calculated by Zipf's law.  Note the generally poorer agreement between the two ranking systems ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
par(mfrow=c(2,2))
par(mar = c(3,3,3,3))

@@

We note that some of these may be generalized; we might consider a
generalized linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for fixed
integer $0<r<n$; the halving system can be generalized to a geometric
distribution; and the winner takes all system can be replaced by
giving equal points to the top $r$ competitors
$(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, for some integer $r<n$.

@@

@@

There is no points system that is the best for all years.  However,
observing that both the likelihood ranking and the points ranking are
random variables in this paradigm suggests a method whereby we can
objectively assess a given points system.  Using sampling techniques
we can repeatedly generate an order table \emph{in silico}, using
estimated driver strengths from the observed tables.  For each of,
say, 1000 such synthetic tables, calculate drivers' maximum likelihood
\citeauthor{plackett1975} strengths, and also their points awarded
according to any given points system.  We then compare rankings
generated by the \citeauthor{plackett1975} strengths and the points
awarded and note the degree of agreement between the two.  This
furnishes an objective assessment of the points system used.

# Numerical results

We now assess the various points systems discussed above, using 1000
_in silico_ trials.

@@

now some summaries:

@@

```{r,label=alottaplotsmeancorrect,echo=FALSE,fig=TRUE}
jj <- cbind(s2016[,1],s2017[,1],s2018[,1],s2019[,1])  # mean

@@

matplot(jj[wanted,],type="n",xlab='points system',ylab='mean correct places',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2,at=1:4)
points(s2016[wanted,1],pch=1)
points(s2017[wanted,1],pch=2)
points(s2018[wanted,1],pch=3)
points(s2019[wanted,1],pch=4)
legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```

```{r,label=alottaplotswinnercorrect,echo=FALSE,fig=TRUE}
jj <- cbind(s2016[,2],s2017[,2],s2018[,2],s2019[,2])  # prob of winner
#matplot(jj,type="n",xlab='points system',ylab='probability of winner correct')
#points(s2016[,2],col=typecol,pch=1)
#points(s2017[,2],col=typecol,pch=2)
#points(s2018[,2],col=typecol,pch=3)
#points(s2019[,2],col=typecol,pch=4)
#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))

matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of winner correct',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2)
points(s2016[wanted,2],pch=1)
points(s2017[wanted,2],pch=2)
points(s2018[wanted,2],pch=3)
points(s2019[wanted,2],pch=4)
legend(x=3.4,y=0.67,lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```

```{r,label=alottaplotstotalrankorder,echo=FALSE,fig=TRUE}
jj <- cbind(s2016[,3],s2017[,3],s2018[,3],s2019[,3])  # total rank order correct
#matplot(jj,type="n",xlab='points system',ylab='probability of total rank order correct')
#points(s2016[,3],col=typecol,pch=1)
#points(s2017[,3],col=typecol,pch=2)
#points(s2018[,3],col=typecol,pch=3)
#points(s2019[,3],col=typecol,pch=4)
#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))

matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of total rank order correct',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2)
points(s2016[wanted,3],pch=1)
points(s2017[wanted,3],pch=2)
points(s2018[wanted,3],pch=3)
points(s2019[wanted,3],pch=4)
legend("topleft",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```

>>>>

\begin{multline}
\mbox{Hamilton}\succ
\mbox{Vettel}\succ
\mbox{Bottas}\succ
\mbox{Raikkonen}\succ\mbox{YES I KNOW}
\mbox{Ricciardo}\succ
\mbox{Verstappen}\succ\\
\mbox{Perez1}\succ
\mbox{Ocon}\succ
\mbox{Sainz}\succ
\mbox{Hulkenberg}\mbox{YES I KNOW}\succ
\mbox{Massa}
\end{multline}

@@

\begin{multline}
\mbox{Hamilton}\succ
\mbox{Vettel}\succ
\mbox{Bottas}\succ
\mbox{Ricciardo}\succ
\mbox{Verstappen}\succ
\mbox{Raikkonen}\succ\mbox{YES I KNOW}\\
\mbox{Perez}\succ
\mbox{Ocon}\succ
\mbox{Massa}\succ
\mbox{Sainz}\succ
\mbox{Hulkenberg}
\end{multline}

@@

```{r,label=likelihoodrankingshower,echo=FALSE}
jj <- mH2017
jj <- sort(jj,decreasing=TRUE)
jj[] <- seq_along(jj)
jj
```

\begin{multline}
\mbox{Hamilton}\succ
\mbox{Vettel}\succ
\mbox{Bottas}\succ
\mbox{Raikkonen}\mbox{YES I KNOW}\succ
\mbox{Ocon}\succ
\mbox{Ricciardo}\succ\\
\mbox{Verstappen}\succ
\mbox{Perez}\succ
\mbox{Massa}\succ
\mbox{Sainz}\succ
\mbox{Hulkenberg}\mbox{YES I KNOW}
\end{multline}

@@

We can plot one ranking against the other\footnote{The historically
correct points awarded to the drivers differs from that calculated
here.  That is for two reasons: firstly, ``fastest lap'' points are
not included here, and also the truncation of the order table to the
first 11 drivers can increase the rank of a driver if non-first-11
drivers are placed}, shown in Figure
\ref{orderingbypointsandlikelihood}.  We thus see a comparison between
the two ordering systems.  Taking 2017 as an example, drivers Hamilton
and Vettel are are respectively first and second according to both
ranking ranking procedures; but R\"{a}ikk\"{o}nen YES I KNOW IT IS
WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
fourth and third, according to points and likelihood respectively.  We
define the \emph{degree of agreement} between the two ranking systems
as the maximal value of $n$ such that places $1,2,\ldots n$ all match.
Thus, from figure \ref{orderingbypointsandlikelihood}, the degree of
agreement between likelihood ranking and points ranking for the years
2016-2019 would be 8,2,6, and 2 respectively.

However, the points system used is essentially arbitrary.  We could
use, for example, Zipf's law to rank the drivers: award one point to
the winner, half a point to second place, one third of a point to
third, and so on; see figure \ref{orderingbypointsandlikelihood_zipf}
in which we see generally poorer agreement between points-based ranks
and likelihood-based ranks, with a degree of agreement of 4,2,2,2 for
the years 2016-2019 respectively.  This might be an indication that
using a Zipf law points allocation is objectively worse than the
current points system.   There are a number of plausible points systems
that might be used:

@@

```{r,label=likevspointszipf,cache=TRUE,echo=FALSE,fig.cap="As for figure \ref{orderingbypointsandlikelihood} \\label{orderingbypointsandlikelihood_zipf} but points ranking calculated by Zipf's law.  Note the generally poorer agreement between the two ranking systems ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
par(mfrow=c(2,2))
par(mar = c(3,3,3,3))

@@

We note that some of these may be generalized; we might consider a
more general linear points system $(r,r-1,r-2,\ldots,3,2,1,0)$ for
fixed integer $0<r<n$; the halving system can be generalized to a
geometric distribution; and the winner takes all system can be
replaced by giving equal points to the top $r$ competitors
$(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, for some integer $r<n$.

@@

It is straightforward to calculate the degree of agreement for the
observed rank table for years 2016-2019, shown in the following table:

@@

It is clear that there is no points system that is the best for all
four years.  However, observing that both the likelihood ranking and
the points ranking are random variables in this paradigm suggests a
method whereby we can objectively assess a given points system.  Using
sampling techniques we can repeatedly generate an order table \emph{in
silico}, using estimated driver strengths from the observed tables.
For each of, say, 1000 such synthetic tables, calculate drivers'
maximum likelihood \citeauthor{plackett1975} strengths, and also their
points awarded according to any given points system.  We then compare
rankings generated by the \citeauthor{plackett1975} strengths and the
points awarded and note the degree of agreement between the two, as
measured by the number of rankings correctly predicted.  This
furnishes an objective assessment of the points system used.

## Numerical results

@@

@@

```{r,label=alottaplotsmeancorrect,echo=FALSE,fig=TRUE,fig.cap="Simulated Formula 1 races, seasons 2016-2019: mean number of agreeing places for each of six points \\label{simulated_mean} systems when compared with a Placket-Luce strength ordering"}
jj <- cbind(s2016[,1],s2017[,1],s2018[,1],s2019[,1])  # mean

@@

matplot(jj[wanted,],type="n",xlab='points system',ylab='mean correct places',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2,at=1:4)
points(s2016[wanted,1],pch=1,type='b')
points(s2017[wanted,1],pch=2,type='b')
points(s2018[wanted,1],pch=3,type='b')
points(s2019[wanted,1],pch=4,type='b')
#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
par(xpd=TRUE)
text(x=6,y=s2016[wanted[6],1],pos=4,"2016")
text(x=6,y=s2017[wanted[6],1],pos=4,"2017")
text(x=6,y=s2018[wanted[6],1],pos=4,"2018")
text(x=6,y=s2019[wanted[6],1],pos=4,"2019")
```

```{r,label=alottaplotswinnercorrect,echo=FALSE,fig=TRUE,fig.cap="Simulated Formula 1 races, seasons 2016-2019: probability of identifying the Plackett-Luce winner for each of six points \\label{simulated_winner_correct} systems"}
jj <- cbind(s2016[,2],s2017[,2],s2018[,2],s2019[,2])  # prob of winner
#matplot(jj,type="n",xlab='points system',ylab='probability of winner correct')
#points(s2016[,2],col=typecol,pch=1)
#points(s2017[,2],col=typecol,pch=2)
#points(s2018[,2],col=typecol,pch=3)
#points(s2019[,2],col=typecol,pch=4)
#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))

matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of winner correct',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2)
points(s2016[wanted,2],pch=1,type='b')
points(s2017[wanted,2],pch=2,type='b')
points(s2018[wanted,2],pch=3,type='b')
points(s2019[wanted,2],pch=4,type='b')
#legend(x=3.4,y=0.67,lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
par(xpd=TRUE)
text(x=6,y=s2016[wanted[6],2],pos=4,"2016")
text(x=6,y=s2017[wanted[6],2],pos=4,"2017")
text(x=6,y=s2018[wanted[6],2],pos=4,"2018")
text(x=6,y=s2019[wanted[6],2],pos=4,"2019")
```

```{r,label=alottaplotstotalrankorder,echo=FALSE,fig=TRUE,fig.cap="Simulated Formula 1 races, seasons 2016-2019: probability of complete agreement between Plackett-Luce ranks and ranks calculated by each of six points \\label{simulated_complete_agreement} systems"}
jj <- cbind(s2016[,3],s2017[,3],s2018[,3],s2019[,3])  # total rank order correct
#matplot(jj,type="n",xlab='points system',ylab='probability of total rank order correct')
#points(s2016[,3],col=typecol,pch=1)
#points(s2017[,3],col=typecol,pch=2)
#points(s2018[,3],col=typecol,pch=3)
#points(s2019[,3],col=typecol,pch=4)
#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))

matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of total rank order correct',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2)
points(s2016[wanted,3],pch=1,type='b')
points(s2017[wanted,3],pch=2,type='b')
points(s2018[wanted,3],pch=3,type='b')
points(s2019[wanted,3],pch=4,type='b')
#legend("topleft",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
par(xpd=TRUE)
text(x=5,y=s2016[wanted[5],3],pos=3,"2016")
text(x=5,y=s2017[wanted[5],3]-0.01,pos=1,"2017")
text(x=5,y=s2018[wanted[5],3],pos=3,"2018")
text(x=5,y=s2019[wanted[5],3],pos=3,"2019")
```

We now assess the six points systems using the methodology outlined
above, using 1000 _in silico_ trials.  For each of the six points
systems, each of the 1000 trials results in a single non-negative
integer: the degree of agreement between the Plackett-Luce ranking and
the rankings according to the points system considered.  There are
three measures that might be used to assess the distribution of degree
of agreement: (1), the mean degree of agreement $\overline{d}$; (2),
the probability of correctly predicting the winner,
$\operatorname{Pr}(d\geqslant 1)$; and (3), the probability of
correctly predicting the complete order statistic
$\operatorname{Pr}(D=n)$.  These summaries are shown for each of the
six points systems in figures \ref{simulated_mean},
\ref{simulated_winner_correct} and \ref{simulated_complete_agreement}
respectively.
